% This is a template for the writeup of your final project.  Note:
% sample-paper.tex is a sample paper created using this template.
% It's a reasonable example of a ten-ish page .tex paper.
%
% To Compile this, do the following:
% > pdflatex sample-paper
% > bibtex sample-paper
% > pdflatex sample-paper
% > pdflatex sample-paper

% definitely use the article class and 11 pt font.
\documentclass[11pt]{article}

% definitely use fullpage package
\usepackage{fullpage}
\usepackage{amsmath,amsthm,amssymb,amsfonts}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning,automata}

\usepackage{mathtools,textcomp,enumerate}
\DeclareMathOperator*{\argmax}{\arg\!\max}


\newcommand{\Tau}{\mathrm{T}}
\newcommand{\E}{\mathrm{E}}
% Include optional other packages here.

% If you have lots of macros and other new command definitions
% consider creating a separate header.tex file.
% If you use this, uncomment the following line:
%\input{header}

% Next, include a title
\title{Hidden Markov Models}

% Put all author information here.
% Since it's just a final project paper, don't feel like you need
%  to include emails or affiliations.
\author{Colin Crowley \& Julian Segert}

% This suppresses the date.
% if you skip it, the date will be automatically be included
\date{}

% starts the document
\begin{document}

% This sets up the title/author/date information
\maketitle

% The abstract is a 1-2 paragraph synopsis of your project and
%  what you accomplished.
\begin{abstract}
Hidden Markov Models (HMMs) are an indispensable tool for modeling and predicting myriad probabilistic processes. Hidden Markov Models involve a chain of state changes and observable emissions at each discrete timestep which are independent of previous timesteps. Because of this independence of events, probabilistic analyses allow us to study many complex systems (or at least approximations of complex systems) in polynomial time.

When using an HMM as a model, there are many values one would like to calculate, or at least estimate. Here, we provide algorithms and analysis for three of the most useful algorithms pertaining to HMMs. The forward-backward calculates the probability distributions of seeing the possible observations at any particular timestep, and the Viterbi algorithm calculates the most likely sequence of observations. Both of these algorithms rely on knowing the model parameters, which cannot always be assumed. For this reason, the Baum-Welch Algorithm was developed to estimate unknown model parameters to within an arbitrary degree of accuracy.
\end{abstract}

% I like to keep each section of a paper in a different .tex file,
%  but you're certainly under no obligation to.
% The \input command essentially cut'n'pastes content from e.g.
%  intro.tex into this .tex file

\input{intro}
\input{analysis.tex}
%\input{prelim}
%\input{firstresult}
%\input{secondresult}
\input{conclusion}
%--------------------------------- bibliography -------------------------
% Thus is a common way to do bibliographies called bibtex
%  write bibtex entries in a .bib file (ours was super.bib)
% Let me know if you have questions on how to this, and/or
%  try to mimic what we do here
\bibliographystyle{plain}
\bibliography{refs}

% Here's how to do appendices
\appendix
%\input{maintechnicalstuff}
%\input{otherstuff}

\end{document}
